Runtime Feedback in a Meta-Tracing JIT for Efficient Dynamic Languages

### 2011-04-15 talk/icooolps2011/paper.tex
some typos and problems I found while reading again

Conceptually, the significant speed-ups that can be achieved with
dynamic compilation depend on feeding into compilation and exploiting
values observed at runtime. In particular, if
there are values which vary very slowly, it is possible to compile multiple
specialized versions of the same code, one for each actual value.  To exploit
the runtime feedback, the implementation code and data structures need to be
structured so that many such slow-varying values are at hand. The hints that
we present precisely allow us to implement such feedback and exploitation in a
meta-tracing context.

@@

The contributions of this paper are:
\begin{itemize}
 \item A hint to turn arbitrary variables into constants in the trace by
 feeding back runtime information into compilation.
 \item A way to annotate operations which the constant folding
 optimization then recognizes and exploits.
 \item A worked-out example of a simple object model of a dynamic language and
 how it can be improved using these hints.
 \item This example also exemplifies general techniques for refactoring code to
 expose likely runtime constants constant folding opportunities.
\end{itemize}

@@

A number of languages have been implemented with PyPy, most importantly a full
Python implementation, but also a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010}.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs have their origin in the Dynamo project, which used one of them for dynamic
assembler optimization \cite{bala_dynamo:_2000}. Later they were used to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

@@

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are turned into machine code. The interpreter is also used when a guard
fails to continue the execution from the failing guard.

@@

Tracing through the execution of an interpreter has many advantages. It makes
the tracer, its optimizers and backends reusable for a variety of languages. The
language semantics do not need to be encoded into the JIT. Instead the tracer
just picks them up from the interpreter. This also means that the JIT by
construction supports the full language.

@@

As the running example of this paper we will use a very simple and bare-bones
object model that just supports classes and instances, without any
inheritance or other advanced features. In the model classes contain methods.
Instances have a class. Instances have their own attributes (or fields). When looking up an
attribute on an instance, the instances attributes are searched. If the
attribute is not found there, the class' methods are searched.

@@

In this section we will describe how to add two hints that allow the
interpreter author to increase the optimization opportunities for constant
folding. If applied correctly these techniques can give really big speedups by
pre-computing parts of what happens at runtime. On the other
hand, if applied incorrectly they might lead to code bloat, thus making the
resulting program actually slower.

@@

The simplest example of constants are literal values, such as \texttt{1}.
However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code on the left. If the fragment is traced with
$x_1$ being \texttt{4}, the trace on the left is produced:

@@

In the trace, the value of $x_1$ is statically known after the guard.
Remember that a guard is a runtime check. The above trace will run to
completion when $x_1$ \texttt{== 4}. If the check fails, execution of the trace is
stopped and the interpreter continues to run.

@@

Promotion is essentially a tool for trace specialization. There are places in
the interpreter where knowing that a value is constant opens a lot of
optimization opportunities, even though it
could have different values in practice. In such a place, promotion can be used. The
typical reason to do that is if there is
a lot of computation depending on the value of that variable.

Let's make this more concrete. If we trace a call to the function on the left, we get the trace on the right:

@@

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let's assume that the value of \texttt{x} in the Python code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:

@@

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the value of $x_1$ as it was during tracing. Thus the runtime value of
\texttt{x} is being made available to the compiler to exploit. The introduced
guard specializes the trace, because it only works if the value of $x_1$ is
\texttt{4}. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the example above. After the guard, the rest of the trace can
assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:

@@

In the previous section we saw a way to turn arbitrary variables into constants. All
foldable operations on these constants can be constant-folded. This works well for
constant folding of simple types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not simple types. The
operations on objects are often not foldable and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

This hint can be used to mark functions as \emph{trace-elidable}. A function is
termed trace-elidable if, during the execution of the program,
successive calls to the function with identical arguments always return the
same result. In addition the function needs to have no side effects or
idempotent side effects\footnote{This property
is less strict than that of a pure function, because it is only about actual
calls during execution. All pure functions are trace-elidable though.}.
From this definition follows that a call to an trace-elidable function with
constant arguments in a trace can be replaced with the result of the call.

@@

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{c}
is an trace-elidable function. To communicate this, there is a \texttt{elidable} decorator.
If the code in \texttt{c} should be constant-folded away, we would change the
class as follows:

@@

    @elidable
    def c(self):
        return self.x * 2 + 1
\end{lstlisting}
\end{minipage}
\vline
\hspace{0.4cm}
\begin{minipage}[t]{0.4\linewidth}
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 
      0xb73984a8)
$v_1$ = c($a_1$)
$v_2$ = $v_1$ + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
\end{minipage}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{c} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{c} function is marked as trace-elidable, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this:
%
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
}

(assuming that the \texttt{x} field's value is \texttt{4}).

On the one hand, the \texttt{elidable} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
function is annotated to be trace-elidable, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation\footnote{The most common use case of the \texttt{elidable}
annotation is indeed to declare the immutability of fields. Because it is so
common, we have special syntactic sugar for it.}. We hope to introduce a
debugging mode which would (slowly) check whether the annotation is applied
incorrectly to mitigate this problem.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Therefore they have to be immutable, which means
that their \texttt{getindex} method is an trace-elidable function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also trace-elidable,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

With this changed instance implementation, the trace we had above changes to the
following that of see Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
memory address of the \texttt{Map} instance that has been promoted. Operations
that can be optimized away are grayed out, their results will be replaced by
fixed values by the constant folding.

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a trace-elidable function and they have constant arguments. That means that \texttt{index1/2/3}
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

@@

Instances were optimized making the assumption that the total number of
different instance layouts is small compared to the number of instances. For classes we
will make an even stronger assumption. We simply assume that it is rare for
classes to change at all. This is not totally reasonable (sometimes classes contain
counters or similar things) but for this simple example it is good
enough.\footnote{There is a more complex variant of class versions that can
accommodate class fields that change a lot better.}

@@

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the \texttt{methods} dictionary changes).
This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a trace-elidable operation.  To help
the JIT to detect this case, we factor it out in a helper method which is
explicitly marked as \texttt{@elidable}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

The index \texttt{0} that is used to read out of the \texttt{storage} array is the result
of the constant-folded \texttt{getindex} call.
The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if the trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{write\_method} method on the class.

@@

Another optimization is that in practice the shape of an instance is correlated
with its class. In our code above, we allow both to vary independently.
Therefore we store the class of an instance on the map in PyPy's Python
interpreter. This means that we get one fewer promotion (and thus one fewer
guard) in the trace, because the class doesn't need to be promoted after the
map has been.

@@

%\subsection{More General Patterns}
%
%The techniques we used above to make instance and class lookups faster are
%applicable in more general cases than the one we developed them for. A more
%abstract view of maps is that of splitting a data-structure into an immutable part (\eg the map)
%and a part that changes (\eg the storage array). All the computation on the
%immutable part is trace-elidable so that only the manipulation of the quick-changing
%part remains in the trace after optimization.
%
%Similarly, versions can be used to constant-fold arbitrary functions of large data
%structures. The version needs to be updated carefully every time the result of
%this function can change. Therefore this is useful only if the data structure is
%expected to change slowly.

@@

For space reasons we cannot perform a full evaluation here, but still want to
present some benchmark numbers. We chose to present just some benchmarks: The
templating engine of the Django web
framework\footnote{\texttt{http://www.djangoproject.com/}}; a Monte-Carlo Go
AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/
disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
Richards benchmark in Python; a Python version of the Telco decimal
benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
pure Python decimal floating point implementation. The results we see in these
benchmarks seem to repeat themselves in other benchmarks using object-oriented
code; for purely numerical algorithms the speedups introduced by the techniques
in this paper are much smaller because they are already fast.

@@

The optimizations give a speedup between 80\% and almost 20 times. The Richards
is a particularly good case for the optimizations as it makes heavy uses of
object-oriented features. Pyflate uses mostly imperative code, so does not
benefit as much. Together with the optimization, PyPy outperforms CPython in
all benchmarks, which is not surprising because CPython is a simple
bytecode-based interpreter.

@@

Yermolovich et. al. describe the use of the Tamarin JavaScript tracing JIT as a
meta-tracer for a Lua interpreter. They compile the normal Lua interpreter in C
to ActionScript bytecode. Again, the interpreter is annotated with some hints
that indicate the main interpreter loop to the tracer.  No further hints are
described in the paper. There is no comparison of their system to the original
Lua VM in C, which makes it hard to judge the effectiveness of the approach.

@@

Partial evaluation \cite{jones_partial_1993} tries to automatically transform
interpreters into compilers using the second futamura projection
\cite{futamura_partial_1999}. Given that classical partial evaluation works
strictly ahead of time, it inherently cannot support runtime feedback.

@@

We already explored promotion in other context, such as earlier versions of
PyPy's JIT as well as a Prolog partial evaluator
\cite{bolz_towards_2009}. Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion was also used in DyC \cite{grant_dyc:_2000}, a
runtime partial evaluator for C. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

@@

The authors would like to thank Peng Wu, David Edelsohn and Laura Creighton for
encouragement, fruitful discussions and feedback during the writing of this paper.

>>>>

Conceptually, the significant speed-ups that can be achieved with
dynamic compilation depend on feeding into compilation values observed at
runtime and exploiting them. In particular, if
there are values which vary very slowly, it is possible to compile multiple
specialized versions of the same code, one for each actual value.  To exploit
the runtime feedback, the implementation code and data structures need to be
structured so that many such slow-varying values are at hand. The hints that
we present precisely allow us to implement such feedback and exploitation in a
meta-tracing context.

@@

The contributions of this paper are:
\begin{itemize}
 \item A hint to turn arbitrary variables into constants in the trace by
 feeding back runtime information into compilation.
 \item A way to annotate operations which the constant folding
 optimization then recognizes and exploits.
 \item A worked-out example of a simple object model of a dynamic language and
 how it can be improved using these hints.
 \item This example also exemplifies general techniques for refactoring code to
 expose constant folding opportunities of likely runtime constants.
\end{itemize}

@@

A number of languages have been implemented with PyPy, most importantly a full
Python implementation, but also a Prolog interpreter
\cite{carl_friedrich_bolz_towards_2010} and some less mature experiments.

@@

A recently popular approach to JIT compilers is that of tracing JITs. Tracing
JITs have their origin in the Dynamo project, which used the technique for dynamic
machine code optimization \cite{bala_dynamo:_2000}. Later they were used to implement
a lightweight JIT for Java \cite{gal_hotpathvm:_2006} and for dynamic languages such as
JavaScript \cite{gal_trace-based_2009}.

@@

To be able to do this recording, VMs with a
tracing JIT typically contain an interpreter. After a user program is
started the interpreter is used; only the most frequently executed paths through the user
program are traced and turned into machine code. The interpreter is also used when a guard
fails to continue the execution from the failing guard.

@@

Tracing through the execution of an interpreter has many advantages. It makes
the tracer, its optimizers and backends reusable for a variety of languages. The
language semantics do not need to be encoded into the JIT. Instead the tracer
just picks them up from the interpreter. This also means that the JIT by
construction supports the full language as correctly as the interpreter.

@@

As the running example of this paper we will use a very simple and bare-bones
object model that just supports classes and instances, without any
inheritance or other advanced features. In the model classes contain methods.
Instances have a class. Instances have their own attributes (or fields). When looking up an
attribute on an instance, the instance's attributes are searched. If the
attribute is not found there, the class' methods are searched.

@@

In this section we will describe two hints that allow the
interpreter author to increase the optimization opportunities for constant
folding. If applied correctly these techniques can give really big speedups by
pre-computing parts of what happens at runtime. On the other
hand, if applied incorrectly they might lead to code bloat, thus making the
resulting program actually slower.

@@

The simplest example of constants are literal values, such as \texttt{1}.
However, the optimizer can statically know the value of a variable even if it
is not a constant in the original source code. For example, consider the
following fragment of RPython code on the left. If the fragment is traced with
\texttt{x} being \texttt{4}, the trace on the right is produced:

@@

A guard is a runtime check. The above trace will run to completion when $x_1$
\texttt{== 4}. If the check fails, execution of the trace is stopped and the
interpreter continues to run. Therefore, the value of $x_1$ is statically known
to be \texttt{4} after the guard.

@@

Promotion is essentially a tool for trace specialization. There are places in
the interpreter where knowing that a value is constant opens a lot of
optimization opportunities, even though it
could have different values in practice. In such a place, promotion can be used. The
typical reason to do that is if there is
a lot of computation depending on the value of one variable.

Let's make this more concrete. If we trace a call to the function (written in
RPython) on the left, we get the trace on the right:

@@

Observe how the first two operations could be constant-folded if the value of
$x_1$ were known. Let's assume that the value of \texttt{x} in the RPython code can vary, but does so
rarely, i.e. only takes a few different values at runtime. If this is the
case, we can add a hint to promote \texttt{x}, like this:

@@

The promotion is turned into a \texttt{guard} operation in the trace. The guard
captures the runtime value of \texttt{x} as it was during tracing, which can
then be exploited by the compiler. The introduced
guard specializes the trace, because it only works if the value of $x_1$ is
\texttt{4}. From the point of view of the
optimizer, this guard is not any different than the one produced by the \texttt{if}
statement in the first example. After the guard, the rest of the trace can
assume that $x_1$ is equal to \texttt{4}, meaning that the optimizer will turn this
trace into:

@@

In the previous section we saw a way to turn arbitrary variables into constants. All
foldable operations on these constants can be constant-folded. This works well for
constant folding of primitive types, e.g. integers. Unfortunately, in the context of an
interpreter for a dynamic
language, most operations actually manipulate objects, not primitive types. The
operations on objects are often not foldable and might even have side-effects. If
one reads a field out of a constant reference to an object this cannot
necessarily be folded away because the object can be mutated. Therefore, another
hint is needed.

This hint can be used to mark functions as \emph{trace-elidable}. A function is
termed trace-elidable if, during the execution of the program,
successive calls to the function with identical arguments always return the
same result. In addition the function needs to have no side effects or
idempotent side effects\footnote{This property
is less strict than that of a pure function, because it is only about actual
calls during execution. All pure functions are trace-elidable though.}.
From this definition follows that a call to an trace-elidable function with
constant arguments in a trace can be replaced with the result of the call seen during tracing.

@@

In this case, adding a promote of \texttt{self} in the \texttt{f} method to get rid of the
computation of the first few operations does not help. Even if $a_1$ is a
constant reference to an object, reading the \texttt{x} field does not necessarily
always yield the same value. To solve this problem, there is another annotation,
which lets the interpreter author communicate invariants to the optimizer. In
this case, she could decide that the \texttt{x} field of instances of \texttt{A} is
immutable, and therefore \texttt{c}
is an trace-elidable function. To communicate this, there is an \texttt{@elidable} decorator.
If the code in \texttt{c} should be constant-folded away, we would change the
class as follows:

@@

    @elidable
    def c(self):
        return self.x * 2 + 1
\end{lstlisting}
\end{minipage}
\vline
\hspace{0.4cm}
\begin{minipage}[t]{0.4\linewidth}
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 
      0xb73984a8)
$v_1$ = A.c($a_1$)
$v_2$ = $v_1$ + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
\end{minipage}

Here, \texttt{0xb73984a8} is the address of the instance of \texttt{A} that was used
during tracing. The call to \texttt{A.c} is not inlined, so that the optimizer
has a chance to see it. Since the \texttt{A.c} method is marked as trace-elidable, and its
argument
is a constant reference, the call will be removed by the optimizer. The final
trace looks like this (assuming that the \texttt{x} field's value is \texttt{4}):
%
{\noop
\begin{lstlisting}[mathescape,basicstyle=\ttfamily]
guard($a_1$ == 0xb73984a8)
$v_2$ = 9 + $val_1$
$a_1$.y = $v_2$
\end{lstlisting}
}

On the one hand, the \texttt{@elidable} annotation is very powerful. It can be
used to constant-fold arbitrary parts of the computation in the interpreter.
However, the annotation also gives the interpreter author ample opportunity to introduce bugs. If a
function is annotated to be trace-elidable, but is not really, the optimizer can produce
subtly wrong code. Therefore, a lot of care has to be taken when using this
annotation\footnote{The most common use case of the \texttt{@elidable}
annotation is indeed to declare the immutability of fields. Because it is so
common, we have special syntactic sugar for it.}. We hope to introduce a
debugging mode which would (slowly) check whether the annotation is applied
incorrectly to mitigate this problem.

@@

In this implementation instances no longer use dictionaries to store their fields. Instead, they have a
reference to a map, which maps field names to indexes into a storage list. The
storage list contains the actual field values. Maps are shared between
different instances, therefore they have to be immutable, which means
that their \texttt{getindex} method is an trace-elidable function. When a new attribute is added
to an instance, a new map needs to be chosen, which is done with the
\texttt{add\_attribute} method on the previous map. This function is also trace-elidable,
because it caches all new instances of \texttt{Map} that it creates, to make
sure that objects with the same layout have the same map, which makes its side
effects idempotent. Now that we have
introduced maps, it is safe to promote the map everywhere, because we assume
that the number of different instance layouts is small.

With this changed instance implementation, the trace we saw in Section~\ref{sub:running} changes to the
that of Figure~\ref{fig:trace2}. There \texttt{0xb74af4a8} is the
memory address of the \texttt{Map} instance that has been promoted. Operations
that can be optimized away are grayed out, their results will be replaced with
fixed values by the constant folding.

The calls to \texttt{Map.getindex} can be optimized away, because they are calls to
a trace-elidable function and they have constant arguments. That means that $index_{1/2/3}$
are constant and the guards on them can be removed. All but the first guard on
the map will be optimized away too, because the map cannot have changed in
between. This trace is already much better than
the original one. Now we are down from five dictionary lookups to just two.

XXX separation of fast and slow-changing parts

@@

Instances were optimized by making the assumption that the total number of
different instance layouts is small compared to the number of instances. For classes we
will make an even stronger assumption. We simply assume that it is rare for
classes to change at all. This is not totally reasonable (sometimes classes contain
counters or similar things) but for this simple example it is good
enough.\footnote{There is a more complex variant of the presented technique that can
accommodate quick-changing class fields a lot better.}

@@

Therefore, we give every class a version object, which is changed every time a
class gets changed (i.e., the \texttt{methods} dictionary changes).
This means that the result of calls to \texttt{methods.get()} for a given \texttt{(name,
version)} pair will always be the same, i.e. it is a trace-elidable operation.  To help
the JIT to detect this case, we factor it out in a helper method \texttt{\_find\_method} which is
marked as \texttt{@elidable}. The refactored \texttt{Class} can
be seen in Figure~\ref{fig:version}

@@

The index \texttt{0} that is used to read out of the \texttt{storage} list is the result
of the constant-folded \texttt{getindex} call.
The constants \texttt{41} and \texttt{17} are the results of the folding of the
\texttt{\_find\_method} calls. This final trace is now very good. It no longer performs any
dictionary lookups. Instead it contains several guards. The first guard
checks that the map is still the same. This guard will fail if the same
code is executed with an instance that has another layout. The second guard
checks that the class of \texttt{inst} is still the same. It will fail if the trace is
executed with an instance of another class. The third guard checks that the
class did not change since the trace was produced. It will fail if somebody
calls the \texttt{write\_method} method on the class.

@@

Another optimization is that in practice the shape of an instance is correlated
with its class. In our code above, we allow both to vary independently.
In PyPy's Python interpreter we store the class of an instance on its map. This
means that we get one fewer promotion and thus one fewer
guard in the trace, because the class doesn't need to be promoted after the
map has been.

@@

%\subsection{More General Patterns}
%
%The techniques we used above to make instance and class lookups faster are
%applicable in more general cases than the one we developed them for. A more
%abstract view of maps is that of splitting a data-structure into an immutable part (\eg the map)
%and a part that changes (\eg the storage list). All the computation on the
%immutable part is trace-elidable so that only the manipulation of the quick-changing
%part remains in the trace after optimization.
%
%Similarly, versions can be used to constant-fold arbitrary functions of large data
%structures. The version needs to be updated carefully every time the result of
%this function can change. Therefore this is useful only if the data structure is
%expected to change slowly.

@@

For space reasons we cannot perform a full evaluation here, but still want to
present some benchmark numbers. We chose to present just some benchmarks: The
templating engine of the Django web
framework\footnote{\texttt{http://www.djangoproject.com/}}; a Monte-Carlo Go
AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/
disco-elegant-python-go-player.html}}; a BZ2 decoder; a port of the classical
Richards benchmark to Python; a Python version of the Telco decimal
benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}}, using a
pure Python decimal floating point implementation. The results we see in these
benchmarks seem to repeat themselves in other benchmarks using object-oriented
code; for purely numerical algorithms the speedups introduced by the techniques
in this paper are much smaller because they are already fast.

@@

The optimizations give a speedup between 80\% and almost 20 times. The Richards benchmark
is a particularly good case for the optimizations as it makes heavy uses of
object-oriented features. Pyflate uses mostly imperative code, so does not
benefit as much. Together with the optimization, PyPy outperforms CPython in
all benchmarks, which is not surprising because CPython is a simple
bytecode-based interpreter.

@@

Yermolovich et. al. \cite{yermolovich_optimization_2009} describe the use of the Tamarin JavaScript tracing JIT as a
meta-tracer for a Lua interpreter. They compile the normal Lua interpreter in C
to ActionScript bytecode. Again, the interpreter is annotated with some hints
that indicate the main interpreter loop to the tracer.  No further hints are
described in the paper. There is no comparison of their system to the original
Lua VM in C, which makes it hard to judge the effectiveness of the approach.

@@

Partial evaluation \cite{jones_partial_1993} tries to automatically transform
interpreters into compilers using the second futamura projection
\cite{futamura_partial_1999}. Given that classical partial evaluation works
strictly ahead of time, it inherently cannot support runtime feedback.
Some partial evaluator work at runtime, such as DyC \cite{grant_dyc:_2000},
which also supports a concept similar to promotion (called dynamic-to-static promotion).

@@

We already explored promotion in other context, such as earlier versions of
PyPy's JIT as well as a Prolog partial evaluator
\cite{bolz_towards_2009}. Promotion is also heavily
used by Psyco \cite{rigo_representation-based_2004} (promotion is called
"unlifting" in this paper) a method-based JIT compiler for Python written by
one of the authors. Promotion is quite similar to
(polymorphic) inline caching and runtime type feedback techniques which were
first used in Smalltalk \cite{deutsch_efficient_1984} and SELF
\cite{hoelzle_optimizing_1991,hoelzle_optimizing_1994} implementations.
Promotion is more general because any information can be cached in line, not
just classes of method receivers.

@@

The authors would like to thank Peng Wu, David Edelsohn and Laura Creighton for
encouragement, fruitful discussions and feedback during the writing of this paper.
This research was partially supported by the BMBF funded project PyJIT (nr. 01QE0913B;
Eureka Eurostars).
